The JuliaHealth Observational Health Research Subecosystem

### 2024-01-06 paper.tex
Resolving additional issues and citation details

% EXPLAIN::MEDIUM Why is this a crucial innovation?
% EXPLAIN::MEDIUM What is meant by a transferable analysis?
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses. \cite{overhage2012validation}
% EXPLAIN::MEDIUM Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries \cite{sachsonOurJourney2023} with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO::EASY Add citation about successful collaborations
% EXPLAIN::EASY What is a network study?
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."\footnotemark

@@

% TODO::MEDIUM This paragraph should be centered on the core technology of the concept of network studies
% TODO::MEDIUM Use this to address the notion of what "network studies" are
% TODO::MEDIUM Make this initially less OHDSI specific; pull from https://jacobzelko.com/05122023034151-network-macarthur
In particular, one collaboration mechanism that OHDSI has been able to operationalize meaningfully while working with observational health data is the notion of "network studies".
These studies, are designed to be transparent and encompasses research endeavors being explored by members of OHDSI comprehensive of relevant documentation, analysis code, results, and study protocols describing the study's scope and objectives.
OHDSI had its inception in the development of the Observational Medical Outcomes Partnership (OMOP) that was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
OHDSI was developed to focus on leveraging observational health data stored within the OMOP CDM.
Since current observational health data sources are aggregated at different levels and cannot be generalized to the overall population, the data is often incomplete, inaccurate, and prone to issues of bias, error handling, and multidimensional complexity. 
The goal of OHDSI is to produce reproducible, robust, and reliable evidence by developing standardized data formats and research protocol designs to enable large-scale collaborative research and improve the transparency and scalability of observational health research. 
% TODO::MEDIUM Carry HADES discussion into the paragraph talking about research tooling
One core innovation brought about by OHDSI in the space of network science is how, given the HADES tools and procedures established by OHDSI, 
% TODO::EASY Make this more general such that it flows into the next paragraph on research tooling better
a research group can develop their analysis on their own OMOP CDM instance and then bundle together their study.
% EXPLAIN::EASY What is a study package?
% EXPLAIN::EASY What data assets?
Once this packaging is done, they can pass along the study package to another site to collaborate on the same analysis but on different data assets.
% EXPLAIN::MEDIUM What is a collaborator network?
Data confidentiality is preserved in this approach as data from individual participating sites remains exclusive to each site, with only aggregate results shared across a given collaborator network.
As such, no risk for re-identification comes about through this network collaboration and has ushered in the possibility for similar research approaches to be taken across the globe. \cite{ohdsi2019book}

@@

% TODO::EASY Segue this better from phenotype definitions
One of the premier tools within the OHDSI HADES ecosystem is the ATLAS Shiny application which was developed to develop patient phenotype definitions within observational health research contexts (amongst other use cases).
These phenotype definitions have varying levels of complexity where they can be very simple (such as finding all patients with one health condition or diagnosis) or grow to be extremely complex (defining controlled, uncontrolled, and indeterminate patient cohorts based on medications, lab results, etc.)
% TODO::EASY Create footnote about schema
% TODO::EASY Cite reference schema from GitHub in footnote; 
% EXPLAIN::EASY Better introduce what SQL is
The computable phenotype definitions that ATLAS generates based on phenotype definitions are in a variety of serialization formats such as JSON or prepared SQL expressions available in multiple SQL flavors.\cite{madrilATLASJSONSchema2022}

% TODO::MEDIUM Reframe this to more say how ATLAS can be used to build cohorts
% TODO::EASY Explain how this package accomplishes similar tasks
The Julia package, OHDSICohortExpressions.jl, exists specifically to ingest these serialized expressions of computable phenotype definitions and create patient cohorts that match the given criterion in the definition. \cite{evansOHDSICohortExpressionsJl2023}
% TODO::MEDIUM Turn all items into complete thoughts in list
How this tool works is to:

@@

\item read and verify the serialized computable phenotype definition

\item reinterpret the serialization into the FunSQL.jl Domain Specific Language

\item return a prepared SQL statement against a given SQL database that queries an OMOP CDM database for the desired patient population(s) of interest.

@@

% TODO::EASY Reword this sentence
Although OHDSICohortExpressions.jl is sufficient to translate existing phenotype definitions generated in ATLAS to executable queries, oftentimes, there is a tricky problem of how to best iterate on phenotype definitions. \cite{zelkoDevelopingRobustComputable2023}
% EXPLAIN:: Better contextualize why this is important -- why not just create a new phenotype definition?
Often, one has to take an existing definition and either revise to a more narrow cohort or write manual code to build more niche cohorts after an initial definition is created.
% EXPLAIN::MEDIUM How so?
% TODO::MEDIUM Review this sentence; maybe cut it
This refactoring of a phenotype definition can lead to conflating ideas within conversation between expert clinical reviewers, analysts, and health researchers.
% TODO::EASY Move this to the previous section on phenotype definitions
This broadly can be the notions of what we are exploring (the phenotype definition; often the responsibility of a clinical reviewer), what is possible to compute upon (the computable phenotype definition; often the responsibility of an analyst), and the goals of the overall study.

@@

% TODO::HARD Review if this subsection is needed
\subsection{Bridging between OHDSI and JuliaHealth Communities}

Briding between tools used in the OHDSI and JuliaHealth communities is now exceptionally straightforward.

% \includegraphics{./atlas_gui.png}

@@

% TODO::HARD Maybe remove package imports and mention them within the caption instead
\begin{listing}[!ht]
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
model = oce.Model(cdm_version = v"5.3.1", 
              cdm_schema = "main",
              vocabulary_schema = "main", 
              results_schema = "main",
              target_schema = "main", 
              target_table = "cohort");

>>>>

% EXPLAIN::MEDIUM Why is this a crucial innovation?
% EXPLAIN::MEDIUM What is meant by a transferable analysis?
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses. \cite{overhage2012validation}
% EXPLAIN::MEDIUM Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries \cite{sachsonOurJourney2023} with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% EXPLAIN::EASY What is a network study?
This has led to successful collaboration relationships \cite{roseFindingAnswersBig2008} across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."\footnotemark

@@

% TODO::MEDIUM This paragraph should be centered on the core technology of the concept of network studies
% TODO::MEDIUM Use this to address the notion of what "network studies" are
% TODO::MEDIUM Make this initially less OHDSI specific; pull from https://jacobzelko.com/05122023034151-network-macarthur
In particular, one collaboration mechanism that OHDSI has been able to operationalize meaningfully while working with observational health data is the notion of "network studies".
These studies, are designed to be transparent and encompasses research endeavors being explored by members of OHDSI comprehensive of relevant documentation, analysis code, results, and study protocols describing the study's scope and objectives.
OHDSI had its inception in the development of the Observational Medical Outcomes Partnership (OMOP) that was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
OHDSI was developed to focus on leveraging observational health data stored within the OMOP CDM.
Since current observational health data sources are aggregated at different levels and cannot be generalized to the overall population, the data is often incomplete, inaccurate, and prone to issues of bias, error handling, and multidimensional complexity. 
The goal of OHDSI is to produce reproducible, robust, and reliable evidence by developing standardized data formats and research protocol designs to enable large-scale collaborative research and improve the transparency and scalability of observational health research. 
% TODO::MEDIUM Carry HADES discussion into the paragraph talking about research tooling
One core innovation brought about by OHDSI in the space of network science is how, given the HADES tools and procedures established by OHDSI, 
% TODO::EASY Make this more general such that it flows into the next paragraph on research tooling better
a research group can develop their analysis on their own OMOP CDM instance and then bundle together their study.
% EXPLAIN::EASY What is a study package?
Once this packaging is done, they can pass along the study package to another site to collaborate on the same analysis but execute the study package on their unique OMOP CDM formatted dataset.
% EXPLAIN::MEDIUM What is a collaborator network?
Data confidentiality is preserved in this approach as data from individual participating sites remains exclusive to each site, with only aggregate results shared across a given collaborator network.
As such, no risk for re-identification comes about through this network collaboration and has ushered in the possibility for similar research approaches to be taken across the globe. \cite{ohdsi2019book}

@@

% TODO::EASY Segue this better from phenotype definitions
One of the premier tools within the OHDSI HADES ecosystem is the ATLAS Shiny application which was developed to develop patient phenotype definitions within observational health research contexts (amongst other use cases).
These phenotype definitions have varying levels of complexity where they can be very simple (such as finding all patients with one health condition or diagnosis) or grow to be extremely complex (defining controlled, uncontrolled, and indeterminate patient cohorts based on medications, lab results, etc.)
The computable phenotype definitions that ATLAS generates based on phenotype definitions are in a variety of serialization formats such as JSON schemas or prepared SQL expressions available in multiple SQL flavors.\footnotemark

\footnotetext{Although the JSON schema is not made formally available within OHDSI, a community made copy of the schema can be found here: \cite{madrilATLASJSONSchema2022}}

% TODO::MEDIUM Reframe this to more say how ATLAS can be used to build cohorts
% TODO::EASY Explain how this package accomplishes similar tasks
The Julia package, OHDSICohortExpressions.jl, exists specifically to ingest these serialized expressions of computable phenotype definitions and create patient cohorts that match the given criterion in the definition. \cite{evansOHDSICohortExpressionsJl2023}
How this tool works is to:

@@

\item Read and verify serialized computable phenotype definitions.

\item Reinterpret serialization into the FunSQL.jl Domain Specific Language.

\item Return a SQL statement for an OMOP CDM database to generate a given cohort.

@@

% TODO::EASY Reword this sentence
Although OHDSICohortExpressions.jl is sufficient to translate existing phenotype definitions generated in ATLAS to executable queries, oftentimes, there is a tricky problem of how to best iterate on phenotype definitions. \cite{zelkoDevelopingRobustComputable2023}
% EXPLAIN:: Better contextualize why this is important -- why not just create a new phenotype definition?
Often, one has to take an existing definition and either revise to a more narrow cohort or write manual code to build more niche cohorts after an initial definition is created.
% TODO::EASY Move this to the previous section on phenotype definitions
This broadly can be the notions of what we are exploring (the phenotype definition; often the responsibility of a clinical reviewer), what is possible to compute upon (the computable phenotype definition; often the responsibility of an analyst), and the goals of the overall study.

@@

@@

\begin{listing}[!ht]
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
model = oce.Model(cdm_version = v"5.3.1", 
              cdm_schema = "main",
              vocabulary_schema = "main", 
              results_schema = "main",
              target_schema = "main", 
              target_table = "cohort");
